---
title: "Multivariate Data Analysis Assignment 3"
output:
  html_document: default
  bookdown::pdf_document2:
    extra_dependencies: ["booktabs", "float"]
    toc: false
    number_sections: false
classoption: twocolumn
---

\raggedbottom

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, results = 'hide', message = FALSE, warning = FALSE,
                      fig.width = 3.2, fig.height = 2.2, fig.pos = 'H')

```


```{r Source and Libraries}
source(file = "./R_SCRIPTS/Functions.R")
seed = seed

library(tidyverse)

D <- load_data()
```

## Introduction

In this report, we are asked to analyse a dataset and present our findings.

The data that we analyse for this report is mRNA gene expression data, from 60 
samples and across 2690 genes from a hypothetical (simulated) study.

The aims of this simulated study are to analyse the effects of two treatments
on a cell culture. The 60 samples are split across two covariates: time (6 or 
12 hours) and treatment (lisinopril, amlodipine, or control).

Our workflow first presents high-level qualities of the data. We move onto 
Principal Components Analysis (PCA) and Cluster Analysis, finding meaningful
partitions of the data. We fit a Linear Discriminant Rule (LDR) to the 
data, and complete the analysis with a brief application of graph-based
clustering to the same data. 


## Part 1: Exploratory Data Analysis



```{r fig-histogram, fig.cap="Histogram Displaying Logarithm of Gene Expression"}

# HISTOGRAM OF LOG GENE EXPRESSION IN A SAMPLE / FOR A PARTICULAR GENE
ggplot() +
  geom_histogram(mapping = aes(x = D$data[1,]), bins = 40, fill = "steelblue", 
                 colour = "white") +
  theme_bw() +
  labs(x = "log Gene Expression",
       y = "Frequency",
       title = "Sa6C1: log Gene Expression") +
  theme(plot.title = element_text(size = 8))

```



```{r fig-boxplot, fig.cap="Boxplot Displaying Distribution of Log Gene Expressions, Split by Covariates"}

# BOXPLOT OF LOG GENE EXPRESSION IN A SAMPLE / FOR A PARTICULAR GENE
selection = c("sample", "times", "treatments")

long_data <- as.data.frame(D$data) |> 
  add_column("times" = D$times, 
             "treatments" = D$treatments) |> 
  rownames_to_column(var = "sample") |> 
  pivot_longer(cols = !any_of((selection)),
               names_to = "gene",
               values_to = "expression")

ggplot(data = long_data, aes(x = treatments, y = expression, fill = treatments,
                             group = sample)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4) +
  facet_wrap(~times, labeller = label_both) +
  scale_fill_manual(values = treatment_colours, guide = "none") +
  theme_bw() +
  labs(x = "Treatments",
       y = "Expression") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Figures \@ref(fig:fig-histogram) and \@ref(fig:fig-boxplot) display the
distributions of gene expression over our 60 samples. We see that the 
expression resembles a unimodal Gaussian distribution with its centre at around
6. 

From Figure \@ref(fig:fig-boxplot), we can see that there are no clear outliers
in the data. All subsets on condition appear normally distributed around the
same centre as the data on aggregate. 


```{r fig-timecourse, fig.cap="Time Course for Gene 1415670_at"}

# PLOT A TIME COURSE FOR A GENE IN DIFFERENT TREATMENTS

time_data <- as.data.frame(D$data[, 1]) |> 
  rename(expression = 1) |> 
  add_column("times" = D$times,
             "treatments" = D$treatments)

ggplot(data = time_data) +
  geom_jitter(mapping = aes(x = times, y = expression, colour = treatments), width = 0.1) +
  geom_line(data = time_data |> 
              group_by(treatments, times) |> 
              summarise(expression = mean(expression), .groups = "drop"),
            mapping = aes(x = times, y = expression, colour = treatments, group = treatments),
            linewidth = 0.8) +
  scale_colour_manual(values = treatment_colours) +
  labs(x = "Time", y = "log Expression", colour = "Treatment") +
  theme_bw()

```

Here in Figure \@ref(fig:fig-timecourse) we examine the differences between 
those samples sequenced at time = 6 and those at time = 12. We can see a 
decrease in expression in the case of the control and lisinopril treatments, 
contrasted with an increase in that of amlodipine. 

## Part 2: Prinicpal Components Analysis

```{r fig-cumulative-variation, fig.cap="Plot of Variation Explained per PC"}

# Run prcomp
PC_hdata <- prcomp(D$data, center = TRUE, scale. = FALSE)

# Compute and plot cumulative variation explained by PC
var_pc <- PC_hdata$sdev^2
tot_var_pc <- sum(PC_hdata$sdev^2)
cumulative_var_pc <- cumsum(var_pc / tot_var_pc)

ggplot() +
  geom_vline(xintercept = 5, colour = "tomato", linewidth = 0.8) +
  geom_point(mapping = aes(x = 1:60, y = cumulative_var_pc)) +
  scale_x_continuous(breaks = function(x) c(scales::breaks_pretty()(x), 5)) +
  theme_bw() +
  labs(x = "PC Index",
       y = "Prop. Variation Explained",
       title = "Variation Explained by PC") +
  theme(plot.title = element_text(size = 8))

```

We choose a value of 5 principal components for downstream
analysis. Here we have used an elbow plot method, but can be confident in our 
low-dimensional representation as these first 5 PCs explain 
`r round(cumulative_var_pc[5], 3)` of the variation. 

```{r fig-pcplot, fig.cap="Plot of the first two PCs against each other."}

# IDENTIFY WHETHER SAMPLES CAN BE GROUPED BY ANY OF THE FACTORS (treatment,
# time, replicate) AND IN WHICH PC DIRECTIONS
PC_hdata_scores <- as.data.frame(PC_hdata$x) |>
  rownames_to_column("sample") |>
  add_column("times" = D$times,
             "treatments" = D$treatments,
             "replicates" = D$replicates)

ggplot(PC_hdata_scores, aes(x = PC1, y = PC2, colour = treatments, shape = factor(times))) +
  geom_point(size = 3) +
  scale_colour_manual(values = treatment_colours) +
  labs(shape = "Times",
       colour = "Treatments") +
  theme_bw() +
  theme(legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"))

```

Here, we plot our data split by the first two principal components, encoding the 
treatment as colour and the time as shape. It is clear to see that we will be 
able to split by these covariates, so separated is our data across them. The plot
gives confidence that we will be able to describe a partition obtained through
cluster or discriminant analysis in terms of these covariates.

Specifically, it looks as if amlodipine samples are split from the rest in the 
first PC direction; and that the times are split along the second PC. We seek
to confirm this more rigorously in the following steps. 

## Part 3: Cluster Analysis

```{r More Libraries}

library(cluster) # kmeans, hierarchical, silhouette
library(factoextra) # hierarchical
library(mclust) # ari
library(fpc) # calinhara

```

In cluster analysis, we look for meaningful partitions of our data. The two
methods we have elected for comparison are KMeans and Hierarchical clustering. 

```{r fig-kmeansplot, fig.cap="Plot of the Partition Obtained by KMeans", fig.keep='last'}

# =====================================================
# KMEANS
# =====================================================

embeddings <- get_embeddings(npcs, D$data)

K_kmeans <- estimate_k(embeddings, FUN = kmeans, b = 100)

K_clustering <- run_kmeans(embeddings, K = K_kmeans$k_val)

ggplot(data = embeddings) +
  geom_point(mapping = aes(x = PC1, y = PC2, colour = as.factor(K_clustering))) +
  labs(colour = "Cluster Assignment") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal") 

```

We use a Gap statistic method to estimate an optimal K to use for clustering,
the specifics of which are in the appendix. We obtain $K = 6$ as our value 
for both algorithms, running the estimation separately for each. 


```{r fig-hierplot, fig.cap="Plot of the Partition Obtained by Hierarchical", fig.keep='last'}

# =====================================================
# HIERARCHICAL
# =====================================================

embeddings <- get_embeddings(npcs, D$data)

# Choose an optimal K for use with hierarchical clustering, this also being 6. 
K_hier <- estimate_k(embeddings, hcut)

# Choice of method based on agglomerative coefficient from cluster::agnes()
method_choice = "ward"

# Run the clustering.
H_clustering <- run_hierarchical(matrix = embeddings, method = method_choice, 
                                 K = K_hier$k_val)

# Plot the partition.
ggplot(data = embeddings) +
  geom_point(mapping = aes(x = PC1, y = PC2, colour = as.factor(H_clustering$cluster))) +
  labs(colour = "Cluster Assignment") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal") 


```
Above are plots of the partitions obtained from each algorithm. At a glance
they are very similar, and in fact we will show they are the same partition.
Note that here we may see an advantage of Hierarchical clustering in that our
workflow produces a nice way to visualise the partitions at different levels of
the hierarchy, in the form of a dendgrogram. 

```{r fig-dendrogram, fig.cap="Dendrogram from Hierarchical Workflow"}

library(dendextend)

dendrogram <- as.dendrogram(as.hclust(H_clustering$hobject))
dendrogram <- color_branches(dendrogram, k = K_hier$k_val)
labels(dendrogram) <- rep("", length(labels(dendrogram)))

par(mar = c(4, 4, 2, 1))
plot(dendrogram, labels = FALSE, main = "Hierarchical Clustering Dendrogram", xlab = "")

colored_bars(colors = cutree(dendrogram, k = K_hier$k_val), dend = dendrogram, sort_by_labels_order = TRUE)

```


```{r tables, results='markup'}

library(broom) 

results <- rbind(
  tidy(chisq.test(table(K_clustering, D$treatments))) |> mutate(comparison = "KMeans ~ Treatments"),
  tidy(chisq.test(table(K_clustering, D$times)))      |> mutate(comparison = "KMeans ~ Times"),
  tidy(chisq.test(table(H_clustering$cluster, D$treatments))) |> mutate(comparison = "Hierarchical ~ Treatments"),
  tidy(chisq.test(table(H_clustering$cluster, D$times)))      |> mutate(comparison = "Hierarchical ~ Times")
)

results$p.value <- format(results$p.value, scientific = TRUE, digits = 2)
knitr::kable(results[, c("comparison", "statistic", "parameter", "p.value")],
      col.names = c("Comparison", "$\\chi^2$", "df", "p"),
      digits = 0,
      format = "latex",
      longtable = FALSE,
      booktabs = TRUE,
      escape = FALSE,
      caption = "Chi-square tests: clustering vs experimental variables")

```

Here we show the output of several Chi-squared tests for the agreement of 
categorical variables, applied to our partition groups and the factor covariates
that we are working with. The interpretation of these is a clear correspondance
between each covariate and the partition. That is to say, we have identified 
six partitions corresponding to each unique combination of our factor covariates. 

```{r tab-partitions, results='markup'}
# TABLE COMPARISON OF BOTH PARTITIONS
knitr::kable(table(H_clustering$cluster, K_clustering),
             format = "latex",
             longtable = FALSE,
             booktabs = TRUE,
             caption = "Comparison of Hierarchical and KMeans Partitions")


# ARANDI COMPARISON OF THE TWO PARTITIONS
HK_ari <- mclust::adjustedRandIndex(H_clustering$cluster, K_clustering)

```

Additionally, we show here that the partitions obtained via KMeans and Hierarchical
are indeed one and the same. The table output of partition agreement is 
completely decoupled. In addition, we output the adjusted Rand Index
measuring agreement between these two partitions. The result is perfect
agreement.

```{r Evaluation of Clustering Performance, results='markup'}

# =====================================================
# EVALUATE CLUSTERING PERFORMANCE
# =====================================================

H_eval <- evaluate_partition(embeddings, H_clustering$cluster, FUN = hcut)

K_eval <- evaluate_partition(embeddings, K_clustering, FUN = kmeans)

H_eval
K_eval

```
These statistics for cluster evaluation confirm our conclusions. A Silhouette 
score of `r round(H_eval$Silhouette, 3)` shows very well separated clusters. 
It should be noted that any slight difference in the Gap statistic for each
partition can be attributed to the randomness inherent in the Gap statistic 
generation process. 


## Part 4: Discriminant Analysis

```{r Discriminant, results='markup'}

library(MASS)

# USING PC SCORES AS VARIABLES, GET A LINEAR DISCRIMINANT ANALYSIS RULE TO
# DISCRIMINATE BETWEEN CELLS AT 12 HOURS (see file)

# Get embeddings and data
D <- load_data()
embeddings <- get_embeddings(npcs, data = D$data)
embeddings$times <- as.factor(D$times)

LDA <- MASS::lda(times ~ PC1 + PC2 + PC3 + PC4 + PC5, data = (embeddings))

predicted <- predict(LDA)$class

# This is our confusion matrix. We see perfect performance from our discriminant
# rule.
knitr::kable(table(D$times, predicted),
             format = "latex",
             booktabs = TRUE,
             caption = "Confusion Matrix: LDA on Time")

```

Here we seek to fit a linear discriminant rule to classify which timestamp
each sample has attached to it, as this is a factor variable with two levels. 

From our discriminant analysis, we see that we have acheived perfect separation.
This tells us that within our low-dimensional representation of the data, there
is no overlap between the groups belonging to each timestamp.

## Part 5: Extension to Graph-Based Clustering

In contemporary mRNA sequencing workflows, alternative cluster analysis methods 
are commonly employed. One such is the Louvain algorithm. It is graph-based,
rather than acting on the expression or similarity matrix directly. 

Here we demonstrate its results on the same data as an extension of the cluster 
analysis in Part 3. 

As is apparent from the name, KNN graph sparsification requires estimation
of an optimal K-value. This is not the number of communities into which to
partition - it is the maximum degree to which we fix each vertex in our graph. 

Whether by fitting a model across a range of values, or by elbow method,
the ways to do this involve evaluating the performance across multiple values
of K and selecting the best, based on some criteria. Commonly, this might be 
the maximum \emph{modularity} obtained when running the Louvain algorithm. 
Modularity is simply the metric that the Louvain algorithm seeks to optimise, 
and is akin to a Calinksi-Harabasz or Silhouette statistic for weighted graphs.

```{r Louvain}

# LIBRARIES
library(tidyverse)
library(igraph)
library(ggraph)

sim_matrix <- 1 / (1 + as.matrix(dist(D$data)))

N <- 5

knn_matrix <- apply(sim_matrix, 1, function(row) {
  threshold <- sort(row, decreasing = TRUE)[N]
  row[row < threshold] <- 0
  return(row)
})

G <- graph_from_adjacency_matrix(knn_matrix, mode = "undirected",
                                 weighted = TRUE)

communities <- cluster_louvain(G, weights = E(G)$weight)
V(G)$community <- membership(communities)
V(G)$name <- rownames(D$data)  

ggraph(G, layout = "fr") +
  geom_edge_link(aes(alpha = weight), show.legend = FALSE) +
  geom_node_point(aes(colour = factor(community)), size = 1) +
  geom_node_text(aes(label = name), repel = TRUE, size = 2) +
  scale_colour_discrete(name = "Community") +
  theme_graph(base_family = "sans")

```
We can see the partition created by our Louvain clustering visualised on our 
KNN graph here. For more complex or larger mRNA expression datasets, we may 
prefer these sorts of algorithms over KMeans or Hierarchical methods. 

## Conclusion

We see clear separation by our factor covariates within this mRNA expression
data. Not only are the groups clear, but they do not overlap at all, which we 
see from a completely decoupled confusion matrix when fitting our linear 
discriminant rule to the data. 

Cluster partitions also separate completely the groups described by our factor
covariates. This gives strength to the assumptions on which KMeans is predicated. 
In particular, the assumption of convex / spherical clusters is supported by 
the performance of our LDA and the clustering itself. 

The data is comprised of distinct groups which are well-described by Treatment
and Time information.









