---
title: "Multivariate Data Analysis Assignment 3"
output:
  html_document: default
  pdf_document: default
---

```{r Source and Libraries}
source(file = "./R_SCRIPTS/Functions.R")

library(tidyverse)

D <- load_data()
```

## Introduction

## Part 1: Exploratory Data Analysis

The data that we analyse for this report is mRNA gene expression data, from 60 
samples and across 2690 genes from a hypothetical (simulated) study.

The aims of this simulated study are to analyse the effects of two treatments
on a cell culture. The 60 samples are split across two covariates: time (6 or 
12 hours) and treatment (lisinopril, amlodipine, or control).

```{r fig-histogram, fig.cap="Histogram Displaying Logarithm of Gene Expression"}

# HISTOGRAM OF LOG GENE EXPRESSION IN A SAMPLE / FOR A PARTICULAR GENE
ggplot() +
  geom_histogram(mapping = aes(x = D$data[1,]), bins = 40, fill = "steelblue", 
                 colour = "white") +
  theme_bw() +
  labs(x = "log Gene Expression",
       y = "Frequency",
       title = "Sa6C1: log Gene Expression Histogram")

```



```{r fig-boxplot, fig.cap="Boxplot Displaying Distribution of Log Gene Expressions, Split by Covariates"}

# BOXPLOT OF LOG GENE EXPRESSION IN A SAMPLE / FOR A PARTICULAR GENE
selection = c("sample", "times", "treatments")

long_data <- as.data.frame(D$data) |> 
  add_column("times" = D$times, 
             "treatments" = D$treatments) |> 
  rownames_to_column(var = "sample") |> 
  pivot_longer(cols = !any_of((selection)),
               names_to = "gene",
               values_to = "expression")

ggplot(data = long_data, aes(x = treatments, y = expression, fill = treatments,
                             group = sample)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4) +
  facet_wrap(~times, labeller = label_both) +
  scale_fill_manual(values = treatment_colours, guide = "none") +
  theme_bw()

```

Figures \@ref(fig:fig-histogram) and \@ref(fig:fig-boxplot) display the
distributions of gene expression over our 60 samples. Note the logarithm of the 
gene expression is given instead of the raw expression value, as the data is 
given to us in log-normalised form. Then, when plotting the data in our
histogram, we see a distribution that resembles a Gaussian one. This we expect. 

From \@ref(fig:fig-boxplot), we can see that the distributions of expression are
not affected to large extent by either of time or treatment. Nor is there much 
variation between replicates; this is expected.


```{r fig-timecourse, fig.cap="Time Course for Gene 1415670_at"}

# PLOT A TIME COURSE FOR A GENE IN DIFFERENT TREATMENTS

time_data <- as.data.frame(D$data[, 1]) |> 
  rename(expression = 1) |> 
  add_column("times" = D$times,
             "treatments" = D$treatments)

ggplot(data = time_data) +
  geom_jitter(mapping = aes(x = times, y = expression, colour = treatments), width = 0.1) +
  geom_line(data = time_data |> 
              group_by(treatments, times) |> 
              summarise(expression = mean(expression), .groups = "drop"),
            mapping = aes(x = times, y = expression, colour = treatments, group = treatments),
            linewidth = 0.8) +
  scale_colour_manual(values = treatment_colours) +
  labs(x = "Time", y = "log Expression", colour = "Treatment") +
  theme_bw()

```

Here in Figure \@ref(fig:fig-timecourse) we examine the differences between 
those samples sequenced at time 6 and those at time 12. We can see a decrease in 
expression in the case of the control and lisinopril treatments, contrasted with
an increase in that of amlodipine. 


## Part 2: Prinicpal Components Analysis

```{r fig-cumulative_variation, fig.cap="Plot of Variation Explained per PC"}

# Run prcomp
PC_hdata <- prcomp(D$data, center = TRUE, scale. = FALSE)

# Compute and plot cumulative variation explained by PC
var_pc <- PC_hdata$sdev^2
tot_var_pc <- sum(PC_hdata$sdev^2)
cumulative_var_pc <- cumsum(var_pc / tot_var_pc)

ggplot() +
  geom_vline(xintercept = 5, colour = "tomato", linewidth = 0.8) +
  geom_point(mapping = aes(x = 1:60, y = cumulative_var_pc)) +
  scale_x_continuous(breaks = function(x) c(scales::breaks_pretty()(x), 5)) +
  theme_bw() +
  labs(x = "PC Index",
       y = "Prop. Variation Explained",
       title = "Proportion of Variation Explained by PC")

```

Where we can confidently choose a value of 5 principal components for downstream
analysis. 

```{r fig-pcplot, fig.cap="Plot of the first two PCs against each other."}

# IDENTIFY WHETHER SAMPLES CAN BE GROUPED BY ANY OF THE FACTORS (treatment,
# time, replicate) AND IN WHICH PC DIRECTIONS
PC_hdata_scores <- as.data.frame(PC_hdata$x) |>
  rownames_to_column("sample") |>
  add_column("times" = D$times,
             "treatments" = D$treatments,
             "replicates" = D$replicates)

ggplot(PC_hdata_scores, aes(x = PC1, y = PC2, colour = treatments, shape = factor(times))) +
  geom_point(size = 3) +
  scale_colour_manual(values = treatment_colours) +
  labs(shape = "Times",
       colour = "Treatments")
  theme_bw()

```

It is clear to see that we will be able to split by these covariates. The plot
gives confidence that we will be able to describe a partition obtained through
cluster or discriminant analysis in terms of these covariates.



## Part 3: Cluster Analysis

```{r More Libraries}

library(cluster) # kmeans, hierarchical, silhouette
library(factoextra) # hierarchical
library(mclust) # ari
library(fpc) # calinhara

```

```{r fig-kmeansplot, fig.cap="Plot of the Partition Obtained by KMeans"}

# =====================================================
# KMEANS
# =====================================================

embeddings <- get_embeddings(npcs, D$data)

K_kmeans <- estimate_k(embeddings, FUN = kmeans, b = 100)

K_clustering <- run_kmeans(embeddings, K = K_kmeans)

ggplot(data = embeddings) +
  geom_point(mapping = aes(x = PC1, y = PC2, colour = as.factor(K_clustering$cluster))) +
  labs(colour = "Cluster Assignment") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal") 

```
```{r fig-hierplot, fig.cap="Plot of the Partition Obtained by Hierarchical"}

# =====================================================
# HIERARCHICAL
# =====================================================

embeddings <- get_embeddings(npcs, D$data)

# Choose an optimal K for use with hierarchical clustering, this also being 6. 
K_hier <- estimate_k(embeddings, hcut)

# Choice of method based on agglomerative coefficient from cluster::agnes()
method_choice = "ward"

# Run the clustering.
H_clustering <- run_hierarchical(matrix = embeddings, method = method_choice, 
                                 K = K_hier$k_val)

# Plot the partition.
ggplot(data = embeddings) +
  geom_point(mapping = aes(x = PC1, y = PC2, colour = as.factor(H_clustering$cluster))) +
  labs(colour = "Cluster Assignment") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal") 


```

```{r Description of Partitions by Covariates}

# =====================================================
# DISCUSS WHETHER WE CAN DESCRIBE OUR CLUSTERS IN TERMS OF THE COVARIATES: TIME.
# TREATMENT, REPLICATES
# =====================================================

# This is adding the partitions onto the df, clustering and then colouring by 
# each of our covariates. 

# KMeans first
table(K_clustering$cluster, D$treatments)
table(K_clustering$cluster, D$times)

chisq.test(table(K_clustering$cluster, D$treatments))
chisq.test(table(K_clustering$cluster, D$times))

# Hierarchical next - need to cuttree first.
table(H_clustering$cluster, D$treatments)
table(H_clustering$cluster, D$times)

chisq.test(table(H_clustering$cluster, D$treatments))
chisq.test(table(H_clustering$cluster, D$times))

# Wow, these are really similar! Are they the same partition?
table(H_clustering$cluster, K_clustering$cluster)
# Yes, the table is decoupled. To really force the point:
HK_ari <- mclust::adjustedRandIndex(H_clustering$cluster, K_clustering$cluster)

```



```{r Evaluation of Clustering Performance}

# =====================================================
# EVALUATE CLUSTERING PERFORMANCE
# =====================================================

H_eval <- evaluate_partition(embeddings, H_clustering$cluster, FUN = hcut)

K_eval <- evaluate_partition(embeddings, K_clustering$cluster, FUN = kmeans)

H_eval
K_eval

```

## Part 4: Discriminant Analysis



## Part 5: Extension to Graph-Based Clustering

## Bibliography