---
title: "Multivariate Data Analysis Assignment 3"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, results = 'hide', message = FALSE, warning = FALSE)

```


```{r Source and Libraries}
source(file = "./R_SCRIPTS/Functions.R")

library(tidyverse)

D <- load_data()
```

## Introduction

In this report, we are asked to analyse a dataset and present our findings.

The data that we analyse for this report is mRNA gene expression data, from 60 
samples and across 2690 genes from a hypothetical (simulated) study.

The aims of this simulated study are to analyse the effects of two treatments
on a cell culture. The 60 samples are split across two covariates: time (6 or 
12 hours) and treatment (lisinopril, amlodipine, or control).

Our workflow first presents high-level qualities of the data. We move onto 
Principal Components Analysis (PCA) and Cluster Analysis, finding meaningful
partitions of the data. We fit a Linear Discriminant Rule (LDR) to the 
data, and complete the analysis with a brief application of graph-based
clustering to the same data. 


## Part 1: Exploratory Data Analysis



```{r fig-histogram, fig.cap="Histogram Displaying Logarithm of Gene Expression"}

# HISTOGRAM OF LOG GENE EXPRESSION IN A SAMPLE / FOR A PARTICULAR GENE
ggplot() +
  geom_histogram(mapping = aes(x = D$data[1,]), bins = 40, fill = "steelblue", 
                 colour = "white") +
  theme_bw() +
  labs(x = "log Gene Expression",
       y = "Frequency",
       title = "Sa6C1: log Gene Expression Histogram")

```



```{r fig-boxplot, fig.cap="Boxplot Displaying Distribution of Log Gene Expressions, Split by Covariates"}

# BOXPLOT OF LOG GENE EXPRESSION IN A SAMPLE / FOR A PARTICULAR GENE
selection = c("sample", "times", "treatments")

long_data <- as.data.frame(D$data) |> 
  add_column("times" = D$times, 
             "treatments" = D$treatments) |> 
  rownames_to_column(var = "sample") |> 
  pivot_longer(cols = !any_of((selection)),
               names_to = "gene",
               values_to = "expression")

ggplot(data = long_data, aes(x = treatments, y = expression, fill = treatments,
                             group = sample)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4) +
  facet_wrap(~times, labeller = label_both) +
  scale_fill_manual(values = treatment_colours, guide = "none") +
  theme_bw()

```

Figures \@ref(fig:fig-histogram) and \@ref(fig:fig-boxplot) display the
distributions of gene expression over our 60 samples. Note the logarithm of the 
gene expression is given instead of the raw expression value, as the data is 
given to us in log-normalised form. Then, when plotting the data in our
histogram, we see a distribution that resembles a Gaussian one. This we expect. 

From \@ref(fig:fig-boxplot), we can see that the distributions of expression are
not affected to large extent by either of time or treatment. Nor is there much 
variation between replicates.


```{r fig-timecourse, fig.cap="Time Course for Gene 1415670_at"}

# PLOT A TIME COURSE FOR A GENE IN DIFFERENT TREATMENTS

time_data <- as.data.frame(D$data[, 1]) |> 
  rename(expression = 1) |> 
  add_column("times" = D$times,
             "treatments" = D$treatments)

ggplot(data = time_data) +
  geom_jitter(mapping = aes(x = times, y = expression, colour = treatments), width = 0.1) +
  geom_line(data = time_data |> 
              group_by(treatments, times) |> 
              summarise(expression = mean(expression), .groups = "drop"),
            mapping = aes(x = times, y = expression, colour = treatments, group = treatments),
            linewidth = 0.8) +
  scale_colour_manual(values = treatment_colours) +
  labs(x = "Time", y = "log Expression", colour = "Treatment") +
  theme_bw()

```

Here in Figure \@ref(fig:fig-timecourse) we examine the differences between 
those samples sequenced at time 6 and those at time 12. We can see a decrease in 
expression in the case of the control and lisinopril treatments, contrasted with
an increase in that of amlodipine. 

## Part 2: Prinicpal Components Analysis

```{r fig-cumulative_variation, fig.cap="Plot of Variation Explained per PC"}

# Run prcomp
PC_hdata <- prcomp(D$data, center = TRUE, scale. = FALSE)

# Compute and plot cumulative variation explained by PC
var_pc <- PC_hdata$sdev^2
tot_var_pc <- sum(PC_hdata$sdev^2)
cumulative_var_pc <- cumsum(var_pc / tot_var_pc)

ggplot() +
  geom_vline(xintercept = 5, colour = "tomato", linewidth = 0.8) +
  geom_point(mapping = aes(x = 1:60, y = cumulative_var_pc)) +
  scale_x_continuous(breaks = function(x) c(scales::breaks_pretty()(x), 5)) +
  theme_bw() +
  labs(x = "PC Index",
       y = "Prop. Variation Explained",
       title = "Proportion of Variation Explained by PC")

```

We can confidently choose a value of 5 principal components for downstream
analysis. Here we have used an elbow plot method, but can be confident in our 
low-dimensional representation as these first 5 PCs explain `r cumulative_var_pc[5]` 
of the variation. 

```{r fig-pcplot, fig.cap="Plot of the first two PCs against each other."}

# IDENTIFY WHETHER SAMPLES CAN BE GROUPED BY ANY OF THE FACTORS (treatment,
# time, replicate) AND IN WHICH PC DIRECTIONS
PC_hdata_scores <- as.data.frame(PC_hdata$x) |>
  rownames_to_column("sample") |>
  add_column("times" = D$times,
             "treatments" = D$treatments,
             "replicates" = D$replicates)

ggplot(PC_hdata_scores, aes(x = PC1, y = PC2, colour = treatments, shape = factor(times))) +
  geom_point(size = 3) +
  scale_colour_manual(values = treatment_colours) +
  labs(shape = "Times",
       colour = "Treatments") +
  theme_bw()

```

It is clear to see that we will be able to split by these covariates. The plot
gives confidence that we will be able to describe a partition obtained through
cluster or discriminant analysis in terms of these covariates.

Specifically, it looks as if amlodipine samples are split from the rest in the 
first PC direction; and that the times are split along the second PC. 

## Part 3: Cluster Analysis

```{r More Libraries}

library(cluster) # kmeans, hierarchical, silhouette
library(factoextra) # hierarchical
library(mclust) # ari
library(fpc) # calinhara

```

In cluster analysis, we look for meaningful partitions of our data. The two
methods we have elected for comparison are KMeans and Hierarchical clustering. 

```{r fig-kmeansplot, fig.cap="Plot of the Partition Obtained by KMeans"}

# =====================================================
# KMEANS
# =====================================================

embeddings <- get_embeddings(npcs, D$data)

K_kmeans <- estimate_k(embeddings, FUN = kmeans, b = 100)

K_clustering <- run_kmeans(embeddings, K = K_kmeans$k_val)

ggplot(data = embeddings) +
  geom_point(mapping = aes(x = PC1, y = PC2, colour = as.factor(K_clustering))) +
  labs(colour = "Cluster Assignment") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal") 

```



```{r fig-hierplot, fig.cap="Plot of the Partition Obtained by Hierarchical"}

# =====================================================
# HIERARCHICAL
# =====================================================

embeddings <- get_embeddings(npcs, D$data)

# Choose an optimal K for use with hierarchical clustering, this also being 6. 
K_hier <- estimate_k(embeddings, hcut)

# Choice of method based on agglomerative coefficient from cluster::agnes()
method_choice = "ward"

# Run the clustering.
H_clustering <- run_hierarchical(matrix = embeddings, method = method_choice, 
                                 K = K_hier$k_val)

# Plot the partition.
ggplot(data = embeddings) +
  geom_point(mapping = aes(x = PC1, y = PC2, colour = as.factor(H_clustering$cluster))) +
  labs(colour = "Cluster Assignment") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal") 


```



```{r tables, results='markup'}

library(broom) 

results <- rbind(
  tidy(chisq.test(table(K_clustering, D$treatments))) |> mutate(comparison = "KMeans ~ Treatments"),
  tidy(chisq.test(table(K_clustering, D$times)))      |> mutate(comparison = "KMeans ~ Times"),
  tidy(chisq.test(table(H_clustering$cluster, D$treatments))) |> mutate(comparison = "Hierarchical ~ Treatments"),
  tidy(chisq.test(table(H_clustering$cluster, D$times)))      |> mutate(comparison = "Hierarchical ~ Times")
)

knitr::kable(results[, c("comparison", "statistic", "parameter", "p.value")],
      col.names = c("Comparison", "X-squared", "df", "p-value"),
      digits = 25,
      caption = "Chi-square tests: clustering vs experimental variables")

```



```{r Description of Partitions by Covariates, results='markup'}
# KMeans first
chisq.test(table(K_clustering, D$treatments))
chisq.test(table(K_clustering, D$times))

# Hierarchical next
chisq.test(table(H_clustering$cluster, D$treatments))
chisq.test(table(H_clustering$cluster, D$times))

# Wow, these are really similar! Are they the same partition?
knitr::kable(table(H_clustering$cluster, K_clustering),
             caption = "Comparison of Hierarchical and KMeans Partitions")


# Yes, the table is decoupled. To really force the point:
HK_ari <- mclust::adjustedRandIndex(H_clustering$cluster, K_clustering)

```



```{r Evaluation of Clustering Performance, results='markup'}

# =====================================================
# EVALUATE CLUSTERING PERFORMANCE
# =====================================================

H_eval <- evaluate_partition(embeddings, H_clustering$cluster, FUN = hcut)

K_eval <- evaluate_partition(embeddings, K_clustering, FUN = kmeans)

H_eval
K_eval

```

## Part 4: Discriminant Analysis

```{r Discriminant, results='markup'}

library(MASS)

# USING PC SCORES AS VARIABLES, GET A LINEAR DISCRIMINANT ANALYSIS RULE TO 
# DISCRIMINATE BETWEEN CELLS AT 12 HOURS (see file)

# Get embeddings and data
D <- load_data()
embeddings <- get_embeddings(npcs, data = D$data)
embeddings$times <- as.factor(D$times)

LDA <- MASS::lda(times ~ PC1 + PC2 + PC3 + PC4 + PC5, data = (embeddings))

predicted <- predict(LDA)$class

table(D$times, predicted)
chisq.test(table(D$times, predicted))

# This is our confusion matrix. We see perfect performance from our discriminant
# rule. 

```

Here we seek to fit a linear discriminant rule to classify which timestamp
each sample has attached to it, as this is a factor variable with two levels. 

From our discriminant analysis, we see that we have acheived perfect separation.
This tells us that within our low-dimensional representation of the data, there
is no overlap between the groups belonging to each timestamp.

## Part 5: Extension to Graph-Based Clustering

In contemporary mRNA sequencing workflows, alternative cluster analysis methods 
are commonly employed. One such is the Louvain algorithm. It is graph-based,
rather than acting on the expression or similarity matrix directly. 

Here we demonstrate its results on the same data as an extension of the cluster 
analysis in Part 3. 

```{r Louvain}

# LIBRARIES
library(tidyverse)
library(igraph)
library(ggraph)

sim_matrix <- 1 / (1 + as.matrix(dist(D$data)))

N <- 5 # set centrally in a second, but a lot of this needs moved over. 

knn_matrix <- apply(sim_matrix, 1, function(row) {
  threshold <- sort(row, decreasing = TRUE)[N]
  row[row < threshold] <- 0
  return(row)
})

G <- graph_from_adjacency_matrix(knn_matrix, mode = "undirected",
                                 weighted = TRUE)
communities <- cluster_louvain(G, weights = E(G)$weight)

V(G)$community <- membership(communities)

V(G)$name <- rownames(D$data)  

ggraph(G, layout = "fr") +
  geom_edge_link(aes(alpha = weight), show.legend = FALSE) +
  geom_node_point(aes(colour = factor(community)), size = 3) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_colour_discrete(name = "Community") +
  theme_graph()

```





